---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

1. Apply packages

Package *dplyr* is used for data processing, and package *ggplot2* is used for ploting.

```{r message=FALSE,warning=FALSE}
library(dplyr)
library(ggplot2)
```

2. Load the data
```{r}
unzip("activity.zip")
activity <- read.csv("activity.csv")
```

## What is mean total number of steps taken per day?

1. Processing the data

Here a new data.frame is made with two columns: one for date and one for the sum of steps of that day.

```{r}
daily.steps <- activity %>% group_by(date) %>% 
    summarize(sum = sum(steps, na.rm = FALSE))
head(daily.steps)
```

2. Plot the histogram
```{r}
h <- ggplot(daily.steps, aes(x = sum))
h + geom_histogram(binwidth = 1000, aes(fill = ..count..)) +
    labs(list(title = "Daily Activity", 
              x = "Number of steps per day", y = "Count"))
```

3. Calculate the mean and median of steps taken per day
```{r}
mean(daily.steps$sum, na.rm = TRUE)
median(daily.steps$sum, na.rm = TRUE)
```

## What is the average daily activity pattern?

1. Data processing

The first line of the code makes a new data.frame containing the average number of steps of each 5-minute interval. The second line converts the class of time interval from character to time ("POSIXlt" "POSIXt"), which lets the time axis of the following plot more straightforward.

```{r}
avg.activity <- activity %>% group_by(interval) %>% 
    summarize(avg = mean(steps, na.rm = TRUE))
avg.activity$time <- strptime(sprintf("%04d", avg.activity$interval), 
                          format="%H%M")
str(avg.activity)
```

2. Make a time series plot
```{r}
library(scales) # for using scale_x_datetime
t <- ggplot(avg.activity, aes(x = time, y = avg))
t + geom_line(aes(colour = avg)) + 
    scale_x_datetime(labels = date_format("%H:%M")) +
    labs(list(title = "Average Daily Activity Pattern", 
              x = "Time interval", 
              y = "Average number of steps"))
```

3. Find the interval with the maximum value
```{r}
avg.activity$interval[which.max(avg.activity$avg)]
```

## Imputing missing values

1. Calculate the total number of missing values in the dataset
```{r}
sum(is.na(activity$steps))
```

2. Filling in all of the missing values

Here the strategy is to fill all of the missing values in the dataset by the mean for that 5-minute interval. It first add a new column containing the average value of that interval ("avg"), and then replace the steps value by it if the value of steps is NA.

```{r}
new.activity <- activity %>% group_by(interval) %>% 
    mutate(avg = mean(steps, na.rm = TRUE)) 
new.activity$steps[!complete.cases(new.activity)] <-
    new.activity$avg[!complete.cases(new.activity)]
head(new.activity)
```

3. Make a histogram of the total number of steps taken each day
```{r}
new.daily.steps <- new.activity %>% group_by(date) %>% 
    summarize(sum = sum(steps, na.rm = TRUE))

h <- ggplot(new.daily.steps, aes(x = sum))
h + geom_histogram(binwidth = 1000, aes(fill = ..count..)) +
    labs(list(title = "Daily Activity (Replacing Missing Data)", 
              x = "Number of steps per day", y = "Count"))
```

4. Calculate the mean and median total number of steps taken per day
```{r}
mean(new.daily.steps$sum)
median(new.daily.steps$sum)
```

## Are there differences in activity patterns between weekdays and weekends?

1. Create a new factor variable with 2 levels -- "weekday" and "weekend"
```{r}
week.activity <- activity %>% mutate(day = weekdays(as.Date(date))) %>%
    mutate(weekday = ifelse(day %in% c("Sunday", "Saturday"), 
                            "weekend", "weekday"))
week.activity$weekday <- as.factor(week.activity$weekday)
head(week.activity)
```

2. Make a weekday v.s. weekend activity pattern

Note here the class of time must be converted to "POSIXct" "POSIXt" or an error will occur when ploting through *ggplot2*.

```{r}
week.activity <- week.activity %>% group_by(weekday, interval) %>% 
    summarize(avg = mean(steps, na.rm = TRUE))
week.activity$time <- as.POSIXct(
    strptime(sprintf("%04d", week.activity$interval), format="%H%M"))
str(week.activity)
```

3. Make a time series plot
```{r}
t <- ggplot(week.activity, aes(x = time, y = avg))
t + geom_line(aes(colour = avg)) + facet_grid(weekday ~ .) + 
    scale_x_datetime(labels = date_format("%H:%M")) +
    labs(list(title = 
                  "Average Daily Activity Pattern: Weekday vs. Weekend", 
              x = "Time interval", 
              y = "Average number of steps")) 
```